Predicting stability of a tuneable ring laser with an iterative model

### 2020-11-18 Papers/Journal/Tuneable_Lasers.tex
Updated some formatting

\author[1]{B. Metherall}
\author[2,*]{C. S. Bohun}

\affil[1]{Mathematical Institute, University of Oxford, Woodstock Rd, Oxford, UK}
\affil[2]{University of Ontario Institute of Technology, 2000 Simcoe St, Oshawa, ON, Canada}

\affil[*]{Corresponding author: sean.bohun@uoit.ca}

@@

\section{Introduction}
\label{sec:intro}
Tuneable lasers have the ability to vary the frequency of its output by up to about 100 nanometres \cite{bohun2015, burgoyne2010, yamashita2009} and simultaneously lase at all frequencies within this bandwidth in short pulses. This tuneability leads to applications such as optical coherence tomography \cite{bohun2015, burgoyne2014, yamashita2009}, coherent anti-Stokes Raman spectroscopy \cite{burgoyne2014}, deep tissue multi-photon microscopy \cite{chung2017, liu2017}, and diagnostics of ultrafast processes \cite{burgoyne2014, silfvast2004}. Tuneable lasers are usually constructed in a ring with five components: an output coupler, chirped fibre Bragg grating (CFBG), modulator, Er-doped fibre, and a pump laser. A typical tuneable ring laser cavity is depicted in Figure \ref{fig:cavity}. Due to the high power and short duration of the pulses, the Kerr nonlinearity becomes crucial to the dynamics within the cavity.

@@

\subsection{Discrete Models}
\label{sec:discrete}
Unfortunately, the master equation, \eqref{eq:meml}, is not completely representative of the underlying physics within our laser cavity. In the derivation of \eqref{eq:meml} it is assumed each process affects the pulse continuously within the cavity. As highlighted by Figure \ref{fig:cavity}, this can be a poor assumption. Within the cavity, each effect is localized to its corresponding component: almost all of the dispersion happens within the CFBG \cite{agrawal2002}, the pulse is only amplified within the Erbium-doped fibre, etc. Thus, a better model is one where we consider a simplified version of the master equation, \eqref{eq:meml}, for each component. Solving these simplified differential equations yields an algebraic expression for the effect of each individual component. The relations can then be functionally composed to give an iterative map for the effect of one round trip of the cavity.

@@

% Table of parameter values
 \begin{table*}[tbp]
 	\centering
 	\begin{tabular}{lcll}
 		\hline\noalign{\smallskip}
 		Parameter & Symbol & Value & Sources \\
 		\hline\noalign{\smallskip}
 		Absorption of Fibre & $\alpha$ & $10^{-4}$--$0.3\text{ m}^{-1}$  & \cite{burgoyneemail, shtyrina2017, tomlinson1984, usechak2005, yarutkina2013} \\
 		Fibre Dispersion & $\beta_2^f$ & $-50$--$50 \text{ ps}^2/ \text{km}$ & \cite{agrawal2002, agrawal2013, burgoyne2014, litchinitser1997, peng2018, yarutkina2013} \\
 		Fibre Nonlinearity & $\gamma$ & $0.001$--$0.01 \text{ W}^{-1} \text{m}^{-1}$ & \cite{agrawal2013, finot2008, usechak2005, yarutkina2013} \\
 		Grating Dispersion & $\beta_2^g L_D$ & $10$--$2000 \text{ ps}^2$ & \cite{agrawal2002, agrawal2013, burgoyne2014, li1998} \\
 		Length of Cavity & $L_T$ & $10$--$100 \text{ m}$ & \cite{burgoyneemail, peng2018, tamura1996} \\
 		Length of Fibre & $L_f$ & $0.15$--$1 \text{ m}$ & \cite{burgoyneemail} \\
 		Length of Gain Fibre & $L_g$ & $2$--$3 \text{ m}$ & \cite{burgoyne2014, peng2018, shtyrina2017, tamura1993,yarutkina2013} \\
 		Modulation Time & $T_M$ & $15$--$150 \text{ ps}$ & \cite{bohun2015, burgoyneemail, burgoyne2014} \\
 		Reflectivity of Optical Coupler & $R$ & $0.1$--$0.9$ & \cite{burgoyneemail, li1998, peng2018,  tamura1993, tamura1996, yamashita2009} \\
 		Saturation Energy & $\Es$ & $10^3$--$10^4 \text{ pJ}$ & \cite{burgoyneemail, usechak2005, yarutkina2013} \\
 		Small Signal Gain & $g_0$ & $1$--$10 \text{ m}^{-1}$ & \cite{burgoyneemail, yarutkina2013} \\
 		\noalign{\smallskip}\hline
 	\end{tabular}
 	\caption{Range of variation of various parameters.}
 	\label{tab:values}
 \end{table*}

@@

Now that we have the algebraic effect of each section of the cavity, we are ready to compose each map together to give the effect of one round trip of the cavity. Thus, we must now consider the order of the components. As we are most interested in the output of the laser cavity, we shall start with the loss. We then pass the pulse through the CFBG followed by the modulator. The pulse then travels through the Er-doped fibre to be amplified, and finally we consider the effect of the nonlinearity. The nonlinearity must immediately follow the gain---this is the section of the cavity where the pulse is most energetic, and hence, has the largest effect. Note, the permutation of the other components is indeed important as the operators do not, in general, commute---this is in contrast to the models described in Section \ref{sec:discrete}. Moreover, we prefer the loss to follow the gain and nonlinearity as we wish to minimize the length over which the nonlinearity has an effect, and maximize the output power. Functionally we denote one round trip of the cavity by
\begin{align}
	\mathcal{L}(A) = F(G(M(D(L(A))))).
	\label{eq:order}
\end{align}
 The pulse after one complete circuit of the laser cavity is then returned back into the cavity to restart the process. A steady solution to this model is one in which the envelope and chirp are unchanged after traversing every component in the cavity---we are uninterested in the phase. That is, such that $\mathcal{L}(A) = A \textrm{e}^{i \phi}$---for some $\phi \in \mathbb{R}$. An example of the evolution of the envelope during one round trip of the laser cavity can be found in Figure \ref{fig:cavityevo}.

@@

To compute $\mathcal{L}(A)$, consider the pulse
\begin{align}
	A = \sqrt{P} \exp \left( -(1 + iC) \frac{T^2}{2 \sigma^2} \right) \textrm{e}^{i \phi_0},
	\label{eq:A0}
\end{align}
where $P$ is the peak power, $C$ is the chirp, $\sigma^2$ is the variance, and $\phi_0$ is the initial phase. This pulse shape is depicted in Figure \ref{fig:samplegauss}.

@@

In Figure \ref{fig:chirpft} we show the chirp and Fourier transform of the pulse \eqref{eq:nlA0} once it has reached equilibrium for the nominal parameter values given in \eqref{eq:ndparam}. In this nonlinear case, we find that the equilibrium pulse is no longer Gaussian---as evidenced by the Fourier transform. Notice that compared to the linear case, higher frequency modes are introduced due to the nonlinearity. Moreover, we find the pulse is linearly chirped near the peak, but, in the tails, $|T| > 1$, the chirp begins to saturate---consistent with experimental results \cite{chen2008, rothenberg1989b, tomlinson1985}. As the nonlinearity parameter, $b$, is increased, the two lobes in the Fourier transform become more pronounced. However, once a critical point is reached SPM begins to play a more substantial factor, leading to modulation instability. This phenomenon is highlighted in Figure \ref{fig:breakevo}. By increasing the nonlinearity, and decreasing the dispersion we find the pulse begins to `breathe'. During the first two dozen round trips of the cavity, the SPM compounds and becomes too parasitic to the pulse. This in turn induces modulation instability, and degrades the pulse, making the pulse no longer stable or sustainable. On the other hand, having a more moderate value of $b$, the pulse is able to equilibrate even with a less favourable initial pulse (Figure \ref{fig:convevo}). The pulse is able to shed the initial left lobe because of dispersion and modulation, and the central lobe is able to grow due to the gain medium. This allows the pulse to come to equilibrium very quickly. Furthermore, since the intensity of the initial pulse is relatively small, the effect of SPM is negligible during the first few round trips which allows the laser to select preferable modes. These preferable modes then get amplified which stabilizes the pulse. From this, it is clear the initial shape of the pulse is much less important than the initial energy of the pulse, $E_0$, or the strength of the nonlinearity, $b$.

@@

We now turn our attention to characterizing the stability of a laser by exploring the parameter space. We focus on the $s$--$b$ plane in particular since $a$ and $h$ effect only the amplitude of the pulse and do not contribute to the interplay of dispersion and the nonlinearity. Additionally, a conservative initial energy will effect the dynamics much less than a more intense pulse since the nonlinearity will be negligible while coming to equilibrium. To identify the stability of a pulse we examine the relative change in the pulse's envelope between consecutive iterations. We compute this error by
\begin{align}
	\textrm{E} = \frac{\| |A_i| - |A_{i-1}| \|_2}{\| A_{i-1} \|_2},
	\label{eq:error}
\end{align}
where $\| \cdot \|_2$ denotes the $L^2(\mathbb{R})$ norm, and $i$ is chosen to be sufficiently large to guarantee the pulse reaches equilibrium or degrades due to modulation instability---we use $i = 250$ in our experiments (Figure \ref{fig:error}). Note that we take the modulus of the pulse since we are only interested in the evolution of the envelope. We find a very sharp and rich boundary between two regions. In the upper-left region the error is $\bigO{1}$, this is where the nonlinearity is too strong, and the SPM induces modulation instability. In the lower-right region we find the opposite behaviour---the pulse reaches an equilibrium state and is stable. The error is on the order of machine precision, or slightly higher due to approximating a smooth continuous function by discrete points.

@@

Also of interest is the energy of the pulse at equilibrium. We compute the energy using \eqref{eq:energy}, or equivalently 
\begin{align}
	E &= \| A_i \|_2,
	\label{eq:energy2norm}
\end{align}
where, again, $i = 250$ in our experiments. Recall from \eqref{eq:order}, this energy is the immediately before passing through the output coupler, and so the energy of each pulse emitted by the laser is instead $(1 - h^2) \| A_i \|_2$, and $h^2 \| A_i \|_2$ remains in the cavity. We plot \eqref{eq:energy2norm} in Figure \ref{fig:energy}. Unsurprisingly, we find the same sharp boundary as we saw in Figure \ref{fig:error}. In the unstable region (upper-left) the energy is relatively small. Since the pulse is not stable in this region it is not able to foster prominent modes, thus, the intensity is low across the entire pulse. In the stable region (lower-right) we find that the energy smoothly decays as both $b$ and $s$ increase, with the contours being hyperbolic-esque. From this it is clear that a more energetic laser requires weaker dispersion, however, the nonlinearity must be correspondingly low, otherwise modulation instability will destroy the pulse.

@@

Finally, we briefly look at how the order of the components effects the dynamics. In Section \ref{sec:effects} we justified the order of the gain, nonlinearity, and loss, but, just assumed dispersion preceded modulation. Here we consider the effect of swapping the dispersion and modulation components. The energy landscape (Figure \ref{fig:energyswitch}) is as we would expect. In the previous case of having modulation following dispersion, more energy is removed from the pulse tails. Whereas, having modulation first, more of this energy remains in the cavity. This has the effect of making the laser more powerful, which in turn increases the unstable region. Comparing with Figure \ref{fig:energy}, this is what we find. The unstable upper-left region spreads farther rightward, and the energy is larger in the stable region---especially farther away from the origin.

>>>>

\author[1]{Brady Metherall}
\author[2,*]{C. Sean Bohun}

\affil[1]{Mathematical Institute, University of Oxford, Radcliffe Observatory Quarter, Woodstock Rd, Oxford OX2 6GG, UK}
\affil[2]{Facutlty of Science, University of Ontario Institute of Technology, 2000 Simcoe St N, Oshawa, ON L1G 0C5, Canada}

\affil[3]{brady.metherall@maths.ox.ac.uk}
\affil[4]{sean.bohun@uoit.ca}

@@

\section{Introduction}
\label{sec:intro}
Tuneable lasers have the ability to vary the frequency of its output by up to about 100 nanometres \cite{bohun2015, burgoyne2010, yamashita2009} and simultaneously lase at all frequencies within this bandwidth in short pulses. This tuneability leads to applications such as optical coherence tomography \cite{bohun2015, burgoyne2014, yamashita2009}, coherent anti-Stokes Raman spectroscopy \cite{burgoyne2014}, deep tissue multi-photon microscopy \cite{chung2017, liu2017}, and diagnostics of ultrafast processes \cite{burgoyne2014, silfvast2004}. Tuneable lasers are usually constructed in a ring with five components: an output coupler, chirped fibre Bragg grating (CFBG), modulator, Er-doped fibre, and a pump laser. A typical tuneable ring laser cavity is depicted in Fig. \ref{fig:cavity}. Due to the high power and short duration of the pulses, the Kerr nonlinearity becomes crucial to the dynamics within the cavity.

@@

\subsection{Discrete Models}
\label{sec:discrete}
Unfortunately, the master equation, \eqref{eq:meml}, is not completely representative of the underlying physics within our laser cavity. In the derivation of \eqref{eq:meml} it is assumed each process affects the pulse continuously within the cavity. As highlighted by Fig. \ref{fig:cavity}, this can be a poor assumption. Within the cavity, each effect is localized to its corresponding component: almost all of the dispersion happens within the CFBG \cite{agrawal2002}, the pulse is only amplified within the Erbium-doped fibre, etc. Thus, a better model is one where we consider a simplified version of the master equation, \eqref{eq:meml}, for each component. Solving these simplified differential equations yields an algebraic expression for the effect of each individual component. The relations can then be functionally composed to give an iterative map for the effect of one round trip of the cavity.

@@

% Table of parameter values
 \begin{table*}[tbp]
	\centering
	\caption{Range of variation of various parameters.}
 	\label{tab:values}
 	\begin{tabular}{lcll}
 		\hline\noalign{\smallskip}
 		Parameter & Symbol & Value & Sources \\
 		\hline\noalign{\smallskip}
 		Absorption of Fibre & $\alpha$ & $10^{-4}$--$0.3\text{ m}^{-1}$  & \cite{burgoyneemail, shtyrina2017, tomlinson1984, usechak2005, yarutkina2013} \\
 		Fibre Dispersion & $\beta_2^f$ & $-50$--$50 \text{ ps}^2/ \text{km}$ & \cite{agrawal2002, agrawal2013, burgoyne2014, litchinitser1997, peng2018, yarutkina2013} \\
 		Fibre Nonlinearity & $\gamma$ & $0.001$--$0.01 \text{ W}^{-1} \text{m}^{-1}$ & \cite{agrawal2013, finot2008, usechak2005, yarutkina2013} \\
 		Grating Dispersion & $\beta_2^g L_D$ & $10$--$2000 \text{ ps}^2$ & \cite{agrawal2002, agrawal2013, burgoyne2014, li1998} \\
 		Length of Cavity & $L_T$ & $10$--$100 \text{ m}$ & \cite{burgoyneemail, peng2018, tamura1996} \\
 		Length of Fibre & $L_f$ & $0.15$--$1 \text{ m}$ & \cite{burgoyneemail} \\
 		Length of Gain Fibre & $L_g$ & $2$--$3 \text{ m}$ & \cite{burgoyne2014, peng2018, shtyrina2017, tamura1993,yarutkina2013} \\
 		Modulation Time & $T_M$ & $15$--$150 \text{ ps}$ & \cite{bohun2015, burgoyneemail, burgoyne2014} \\
 		Reflectivity of Optical Coupler & $R$ & $0.1$--$0.9$ & \cite{burgoyneemail, li1998, peng2018,  tamura1993, tamura1996, yamashita2009} \\
 		Saturation Energy & $\Es$ & $10^3$--$10^4 \text{ pJ}$ & \cite{burgoyneemail, usechak2005, yarutkina2013} \\
 		Small Signal Gain & $g_0$ & $1$--$10 \text{ m}^{-1}$ & \cite{burgoyneemail, yarutkina2013} \\
 		\noalign{\smallskip}\hline
 	\end{tabular}
 \end{table*}

@@

Now that we have the algebraic effect of each section of the cavity, we are ready to compose each map together to give the effect of one round trip of the cavity. Thus, we must now consider the order of the components. As we are most interested in the output of the laser cavity, we shall start with the loss. We then pass the pulse through the CFBG followed by the modulator. The pulse then travels through the Er-doped fibre to be amplified, and finally we consider the effect of the nonlinearity. The nonlinearity must immediately follow the gain---this is the section of the cavity where the pulse is most energetic, and hence, has the largest effect. Note, the permutation of the other components is indeed important as the operators do not, in general, commute---this is in contrast to the models described in Section \ref{sec:discrete}. Moreover, we prefer the loss to follow the gain and nonlinearity as we wish to minimize the length over which the nonlinearity has an effect, and maximize the output power. Functionally we denote one round trip of the cavity by
\begin{align}
	\mathcal{L}(A) = F(G(M(D(L(A))))).
	\label{eq:order}
\end{align}
 The pulse after one complete circuit of the laser cavity is then returned back into the cavity to restart the process. A steady solution to this model is one in which the envelope and chirp are unchanged after traversing every component in the cavity---we are uninterested in the phase. That is, such that $\mathcal{L}(A) = A \textrm{e}^{i \phi}$---for some $\phi \in \mathbb{R}$. An example of the evolution of the envelope during one round trip of the laser cavity can be found in Fig. \ref{fig:cavityevo}.

@@

To compute $\mathcal{L}(A)$, consider the pulse
\begin{align}
	A = \sqrt{P} \exp \left( -(1 + iC) \frac{T^2}{2 \sigma^2} \right) \textrm{e}^{i \phi_0},
	\label{eq:A0}
\end{align}
where $P$ is the peak power, $C$ is the chirp, $\sigma^2$ is the variance, and $\phi_0$ is the initial phase. This pulse shape is depicted in Fig. \ref{fig:samplegauss}.

@@

In Fig. \ref{fig:chirpft} we show the chirp and Fourier transform of the pulse \eqref{eq:nlA0} once it has reached equilibrium for the nominal parameter values given in \eqref{eq:ndparam}. In this nonlinear case, we find that the equilibrium pulse is no longer Gaussian---as evidenced by the Fourier transform. Notice that compared to the linear case, higher frequency modes are introduced due to the nonlinearity. Moreover, we find the pulse is linearly chirped near the peak, but, in the tails, $|T| > 1$, the chirp begins to saturate---consistent with experimental results \cite{chen2008, rothenberg1989b, tomlinson1985}. As the nonlinearity parameter, $b$, is increased, the two lobes in the Fourier transform become more pronounced. However, once a critical point is reached SPM begins to play a more substantial factor, leading to modulation instability. This phenomenon is highlighted in Fig. \ref{fig:breakevo}. By increasing the nonlinearity, and decreasing the dispersion we find the pulse begins to `breathe'. During the first two dozen round trips of the cavity, the SPM compounds and becomes too parasitic to the pulse. This in turn induces modulation instability, and degrades the pulse, making the pulse no longer stable or sustainable. On the other hand, having a more moderate value of $b$, the pulse is able to equilibrate even with a less favourable initial pulse (Fig. \ref{fig:convevo}). The pulse is able to shed the initial left lobe because of dispersion and modulation, and the central lobe is able to grow due to the gain medium. This allows the pulse to come to equilibrium very quickly. Furthermore, since the intensity of the initial pulse is relatively small, the effect of SPM is negligible during the first few round trips which allows the laser to select preferable modes. These preferable modes then get amplified which stabilizes the pulse. From this, it is clear the initial shape of the pulse is much less important than the initial energy of the pulse, $E_0$, or the strength of the nonlinearity, $b$.

@@

We now turn our attention to characterizing the stability of a laser by exploring the parameter space. We focus on the $s$--$b$ plane in particular since $a$ and $h$ effect only the amplitude of the pulse and do not contribute to the interplay of dispersion and the nonlinearity. Additionally, a conservative initial energy will effect the dynamics much less than a more intense pulse since the nonlinearity will be negligible while coming to equilibrium. To identify the stability of a pulse we examine the relative change in the pulse's envelope between consecutive iterations. We compute this error by
\begin{align}
	\textrm{E} = \frac{\| |A_i| - |A_{i-1}| \|_2}{\| A_{i-1} \|_2},
	\label{eq:error}
\end{align}
where $\| \cdot \|_2$ denotes the $L^2(\mathbb{R})$ norm, and $i$ is chosen to be sufficiently large to guarantee the pulse reaches equilibrium or degrades due to modulation instability---we use $i = 250$ in our experiments (Fig. \ref{fig:error}). Note that we take the modulus of the pulse since we are only interested in the evolution of the envelope. We find a very sharp and rich boundary between two regions. In the upper-left region the error is $\bigO{1}$, this is where the nonlinearity is too strong, and the SPM induces modulation instability. In the lower-right region we find the opposite behaviour---the pulse reaches an equilibrium state and is stable. The error is on the order of machine precision, or slightly higher due to approximating a smooth continuous function by discrete points.

@@

Also of interest is the energy of the pulse at equilibrium. We compute the energy using \eqref{eq:energy}, or equivalently 
\begin{align}
	E &= \| A_i \|_2,
	\label{eq:energy2norm}
\end{align}
where, again, $i = 250$ in our experiments. Recall from \eqref{eq:order}, this energy is the immediately before passing through the output coupler, and so the energy of each pulse emitted by the laser is instead $(1 - h^2) \| A_i \|_2$, and $h^2 \| A_i \|_2$ remains in the cavity. We plot \eqref{eq:energy2norm} in Fig. \ref{fig:energy}. Unsurprisingly, we find the same sharp boundary as we saw in Fig. \ref{fig:error}. In the unstable region (upper-left) the energy is relatively small. Since the pulse is not stable in this region it is not able to foster prominent modes, thus, the intensity is low across the entire pulse. In the stable region (lower-right) we find that the energy smoothly decays as both $b$ and $s$ increase, with the contours being hyperbolic-esque. From this it is clear that a more energetic laser requires weaker dispersion, however, the nonlinearity must be correspondingly low, otherwise modulation instability will destroy the pulse.

@@

Finally, we briefly look at how the order of the components effects the dynamics. In Section \ref{sec:effects} we justified the order of the gain, nonlinearity, and loss, but, just assumed dispersion preceded modulation. Here we consider the effect of swapping the dispersion and modulation components. The energy landscape (Fig. \ref{fig:energyswitch}) is as we would expect. In the previous case of having modulation following dispersion, more energy is removed from the pulse tails. Whereas, having modulation first, more of this energy remains in the cavity. This has the effect of making the laser more powerful, which in turn increases the unstable region. Comparing with Fig. \ref{fig:energy}, this is what we find. The unstable upper-left region spreads farther rightward, and the energy is larger in the stable region---especially farther away from the origin.
